Sequential rank agreement methods for comparison of lists

### 2015-07-25 latex/paper.Rnw
Applicaitons

\subsection{Agreement of predictor ranking}

@@

\subsection{Agreement of predictions}

Here we apply the same statistical classification
method to XX bootstrap samples of the MALOVA data to obtain four
prediction models. From each model we predict the risks of a malignant
patients in the DACOVA study.  Finally we estimate the agreement of
the resulting XX rankings of the DACOVA patients to illustrate the
sensitivity of the corresponding statistical classification method to
pertubations of the learning data.

@@

@@

>>>>

\subsection{Agreement of predictor ranking}
We apply 10 different classification methods to each of the two data
sets. All methods are capable of producing a ranking of the 5000
predictors according to their importance with respect to the
association with the outcome class label. The methods under
consideration can be separated into two groups: marginal methods and
joint methods. The marginal methods treat each of the of 5000
predictors separately in succession while the joint models use all the
predictors simultaneously. The marginal methods are t-test statistics,
univariate logistic regressions, AUC-based ROC analyses and maximum
information content correlations (MIC) statistics. For the t-test
statistics and the logistic regressions we rank the predictors
according to decreasing p-value while for the ROC analysis and MIC
statistics we simply rank according to increasing numerical value. The
joint methods are a Random Forest classifier
\citep{breiman2001random}, logistic Lasso regression
\citep{tibshirani1996regression}, logistic Ridge regression
\cite{segerstedt1992ordinary}, Partial Least Squares Discriminant
Analysis (PLS-DA) \cite{boulesteix2004pls}, Shrinkage Discriminant
Analysis with local false discovery rate control (SDA)
\cite{ahdesmaki2010} and nearest shrunken centroid (PAM)
\cite{tibshirani2003class}. For the Random Forest classifier the
importance of the predictors are ranked according to the Gini index
while for the logistic Lasso and Ridge regressions we order by
absolute magnitude of the estimated coefficients. For the PLS-DA
method the importance of the predictors is based on a weighted sum of
the absolute coefficients where the weights are proportional to the
reduction in the sums of squares across the components, and for PAM
the importance is given by the difference between the class centroids
and the overall centroid. Finally, for the SDA method the ranking of
the predictors is performed using the correlation-adjusted t-scores,
the local false discovery rate (lFDR) is estimated and only the
predictors with $lFDR < 0.8$ are retained. Together with the logistic
Lasso regression and the nearest shrunken centroid methods this leads
to a censored list according to our previous discussion. All tuning
parameters are estimated by repeated 10-fold cross-validation using 20
random splits. We use the implementations in the R packages
\texttt{randomForest} \citep{liaw2002classification}, \texttt{glmnet}
\citep{Friedman:2010aa}, \texttt{caret} \citep{Jed-Wing:2014aa},
\texttt{sda} \citep{Ahdesmaki:2014aa} and \texttt{minerva}
\citep{Filosi:2014aa}.

The agreement of the ranking of the predictors given by the ten
different methods can be compared in different ways using the
sequential rank agreement. One possibility is to consider all $55$
pair-wise agreement functions. This will be useful for identifying
statistical methods that perform similarly on a given data set with
respect to how the order of predictor importance is determined. This
could for example be done by applying a statistical clustering method
to the pair-wise sequential rank agreement curves. Here we take a
simpler approach and consider the agreement of the importance
determined by the different marginal methods and the agreement of the
importance determined by the different joint methods. This leads to
two agreement curves for each data set which we show in figure
\ref{fig:app1}.

@@

From the figure it is seen that generally the agreement between
ranking the predictors using the different methods is better for the
MALOVA data than for the DACOVA data. The shapes of the agreement
curves start at a low level followed by an increase after which they
stabilize at some level. This corresponds to the expected behavior
that the methods agree better with respect to the most important
predictors than with the not so important ones. Also, it is clear that
ranking the predictors by the marginal methods show better overall
agreement than using the joint methods. This is not surprising due to
the high correlation of the spectral intensities across neighboring
mass-to-charge ratios. This leads to co-linearity problems that
especially affects the classification methods that use a sparse
representation of the predictors.

\subsection{Agreement of predicted risk probabilities}
We now consider the case where we predict the patient-specific risks
of having a malignant tumor. We consider this as an example of the
clinical decision problem of deciding the order in which a group of
patients must be treated. This order is given by ranking the patients
according to increasing predicted probability of malignant tumor. We
assess the agreement of estimating the rankings of the same patients
by four different statistical prediction methods using the sequential
rank agreement. The four prediction models are: a Random Forest
classifier, logistic Lasso regression, logistic Ridge regression and
Partial Least Squares Discriminant Analysis (PLSDA).

For each prediction method we estimate a model based on the MALOVA
data and predict the risks of a malignant tumor for the patients in
the DACOVA study. We repeat this procedure on 100 bootstrap samples of
the MALOVA data used to construct the model and then estimate the
agreement of the resulting 100 rankings of the same DACOVA
patients. This serves as an assessment of the sensitivity of the
different statistical prediction methods to perturbations of the
learning data.

For the Random Forest method we grow 500 tree using 22 randomly
selected predictors at each split. The penalty parameter for the Lasso
and Ridge regression models is optimized with respect to the deviance
loss function using internal repeated 10-fold cross-validation on the
learning data using 20 random splits. The number of components for the
PLSDA method is optimized in a similar manner using classification
accuracy as criterion.

Figure \ref{fig:app2} shows the sequential rank agreement for the four
methods. Generally, the ranking of the predictions made by the Random
Forest method appears to be in better agreement across resampled
training data than for any of the other method. The only exception is
for the patient with the overall largest risk of a malignant tumor for
which the logistic Lasso regression method provides a better
agreement. Something about Ridge is surprising etc. etc. etc. that we
now look at in a simulation study.

@@

To further investigate the apparent lack of agreement for the ridge
regression method we consider a small simulation study. We simulate
two data sets each with $120$ observations and $500$ predictors. Each
vector of predictors is drawn independently from a $500$-dimensional
Gaussian distribution with zero mean, unit variance and a correlation
coefficient of $0.9$ in a compound symmetry structure. The 120
observations is split into two groups where the coefficient vector for
the first group is zero and similarly for the second group except that
the first $100$ values are set to one. We then predict the risk
probabilities in a similar manner as before where we repeatedly fit
models on bootstrapped versions on the first data set and use the
models to predict the risk probability of the same observations from
the second data set. Here we only consider the logistic Ridge
regression model but we consider three different ways of optimizing
the penalty parameter. These consists of optimizing with respect to
the total deviance, the prediction accuracy and the area under the ROC
curve. The resulting sequential rank agreement curves from 100
bootstrap iterations are seen in figure \ref{fig:app3}

@@

Redo the figure and comment on the result. Note that none of the loss
functions directly considers the agreement of predicted
probabilities. The AUC is indirectly related in a complicated manner.
